---
title: "Great Plains Fire Zone"
format: 
  html:
    page-layout: full
editor: source

---

```{r}
#| label: set up
#| message: false
#| warning: false
#| include: false

library(htmltools)
library(htmlwidgets)
library(plotly)
library(scales)
library(stringr)
library(tidygraph)
library(tidyverse)



options(scipen = 99999999)

bps_data <- read.csv("data/great_plains/bps_aoi_attributes.csv")


bps_data[bps_data == -9999] <- 0

bps_data <- bps_data[bps_data$BPS_NAME != "Open Water", ]


evt_data <- read_csv(file = "data/great_plains/evt_aoi_attributes.csv")

evc_data <- read.csv("data/great_plains/evc_aoi_attributes.csv") 

evh_data <- read.csv("data/great_plains/evh_aoi_attributes.csv") 


# process fire data to get acres up front

all_annual_fire <-bps_data %>%
  mutate(annual_fire_acres = ((1/FRI_ALLFIR)*ACRES)) %>%
  filter(BPS_NAME != 'Open Water') %>%
  group_by(BPS_NAME) %>%
  summarize(acres = sum(annual_fire_acres)) %>%
  arrange(desc(acres))


all_annual_fire$acres[all_annual_fire$acres == Inf] <- 0


annualFire <- all_annual_fire %>%
   top_n(n = 10, wt = acres) 





fire_data <- bps_data |>
  mutate(annual_fire_acres = ((1/FRI_ALLFIR)*ACRES)) |>
  mutate(annual_no_fire = (ACRES - annual_fire_acres)) |>
  mutate(annual_surface_acres = ((1/FRI_SURFAC)*ACRES)) |>
  mutate(annual_mixed_acres = ((1/FRI_MIXED)*ACRES)) |>
  mutate(annual_replacement_acres = ((1/FRI_REPLAC)*ACRES)) |>
  group_by(BPS_NAME) |>
  summarize(ACRES = sum(ACRES),
            REL_PERCENT = sum(REL_PERCENT),
            annual_fire_acres = sum(annual_fire_acres),
            annual_no_fire = sum(annual_no_fire),
            annual_surface_acres = sum(annual_surface_acres),
            annual_mixed_acres = sum(annual_mixed_acres),
            annual_replacement_acres = sum(annual_replacement_acres),) |>
  arrange(desc(REL_PERCENT)) |>
  top_n(n = 10, wt = REL_PERCENT)

fire_data <- fire_data %>%
  rename(Surface = annual_surface_acres, 
         Mixed = annual_mixed_acres, 
         Replacement = annual_replacement_acres, 
         No_fire = annual_no_fire)

fire_data[fire_data == Inf] <- 0

fire_data_long <- fire_data %>%
  pivot_longer(cols = c(Surface, Mixed, Replacement, No_fire),
               names_to = "fire_type",
               values_to = "fire_acres") %>%
  mutate(fire_type = factor(fire_type, levels = c("Surface", 
                                                  "Mixed", 
                                                  "Replacement", 
                                                  "No_fire")))



```





## Historical Fire Regimes


<br>

::: {.callout-note}
[ **Prior to European colonization the Great Plains fire zone had `r formatC(sum(all_annual_fire$acres), format = "f", big.mark = ",", digits = 0)` acres of fire annually.** ]{style="font-size: 1.5em;"}
:::

<br> <br>

### Historical Annual Fire Amounts and Their Associated Biophysical Settings

In the interactive chart below, hover over the various elements to learn how much of each BpS was on the landscape historically, and how much fire across the 3 main fire types happened annually prior to European Settlement.

```{r}
#| label: mfri and bpss
#| echo: false
#| message: false
#| warning: false
#| fig-height: 12
#| output-height: 12
# Read in and reshape fire data


# Read in snkyPdata
snkyPdata <- fire_data_long |>
  select(c(fire_type, BPS_NAME, fire_acres)) |>
  rename(source = BPS_NAME, target = fire_type, value = fire_acres)

# Convert factor variables to character
snkyPdata <- snkyPdata %>%
  mutate(across(everything(), as.character))

# Replace "No_fire" with "No Fire"
snkyPdata <- snkyPdata %>%
  mutate(across(everything(), ~replace(., . == "No_fire", "No Fire")))

# Define the desired order for fire types (targets), placing No_fire last
target_order <- c('Surface', 'Mixed', 'Replacement', 'No Fire')

# Sort data by target order
snkyPdata <- snkyPdata %>%
  mutate(target = factor(target, levels = target_order)) %>%
  arrange(target)

# Define node names: sources first (unique BPS_NAME), then targets (in order)
all_nodes <- c(unique(snkyPdata$source), target_order)
nodes <- data.frame(name = all_nodes)

# Define colors
sourceColor <- '#bbc4bc'
targetColors <- c('#d95f0e', '#fec44f', '#fff7bc', '#117733')  # matches target_order
nodeColors <- c(rep(sourceColor, length(unique(snkyPdata$source))), targetColors)

# Create source and target IDs
snkyPdata$IDsource <- match(snkyPdata$source, nodes$name) - 1
snkyPdata$IDtarget <- match(snkyPdata$target, nodes$name) - 1

# Generate Sankey diagram
sankey <- plot_ly(
  type = "sankey",
  orientation = "h",
  node = list(
    pad = 15,
    thickness = 20,
    line = list(color = "black", width = 0.5),
    label = nodes$name,
    color = nodeColors,
    hovertemplate = "Acres"
  ),
  link = list(
    source = snkyPdata$IDsource,
    target = snkyPdata$IDtarget,
    value = snkyPdata$value,
    hovertemplate = "Acres"
  ),
  textfont = list(size = 14),
  width = 850,
  height = 800 # taller to help spacing
) %>%
  layout(
    title = "",
    font = list(size = 14),
    margin = list(t = 100, l = 20, r = 20, b = 100)
  )

sankey

```


<br>

### Biophysical Settings descriptions

Below you will find a list of Biophysical Settings (BpS) that is linked to their descriptions.  Note:

* There may be multiple BpS names listed.  If so, this is because your landscape spans multiple LANDFIRE Map Zones and has multiple 'variants' for those BPSs.  
* The names below are followed by a number in parentheses.  The first 5-digit number is the BPS Code.  The following numbers reprent the Map Zones.  For example, "13020_65_66" is ‘Laurentian-Acadian Northern Hardwoods forest’ BpS (13020) in Map Zones 65 and 66.
* The map below will help orient you to the Map Zones for your landscape.

<br>


```{r}
#| label: mz map
#| echo: false
#| message: false
#| warning: false
library(sf)
library(tmap)

tmap_mode("view")


# Replace 'path_to_map_zones_shapefile' and 'path_to_additional_shapefile' with the actual paths to your shapefiles
map_zones <- st_read("data/conus_mapzones_102611.shp", quiet = TRUE)
additional_shape <- st_read("data/great_plains/great_plains_tracts.shp", quiet = TRUE)


map_zones <- st_transform(map_zones, crs = 4326)
additional_shape <- st_transform(additional_shape, crs = 4326)


# Create the tmap


tm_shape(map_zones) +
  tm_borders(col = "black", lwd = 2) +
  tm_text("ZONE_NUM", size = 1, col = "black", fontface = "bold") +
  tm_shape(additional_shape) +
  tm_borders(col = "green", lwd = 2) +
  tm_fill(fill = "green", fill_alpha = 0.5) +
  tm_view(bbox = st_bbox(additional_shape)) +
  tm_layout(legend.show = FALSE)
```

<br>



**Clicking link will initiate a download of a Microsoft Word file.  If you have any issues with this please contact [LANDFIRE](mailto:landfire@tnc.org){target="blank"}.**


```{r}
#| label: get bps docs
#| echo: false
#| message: false
#| warning: false

# want list of BpS names to match chart
# should have variants with their percents
# should be in order by percent with the correct percent




# Get unique model codes for the top 10 BPS names
unique_model_codes <- unique(bps_data$BPS_MODEL[bps_data$BPS_NAME %in% fire_data$BPS_NAME])

# Filter the dataframe to include only rows with unique model codes and top 10 BPS names
bps_urls <- bps_data %>% 
  filter(BPS_MODEL %in% unique_model_codes) %>% 
  distinct(BPS_MODEL, .keep_all = TRUE) %>%
  select(c(BPS_NAME, BPS_MODEL))


# Merge top_bps with bps_urls to get REL_PERCENT
bps_urls <- bps_urls %>%
  left_join(bps_data |>
              select(BPS_MODEL, REL_PERCENT), 
            by = "BPS_MODEL")

# URL template
url_template <- "https://github.com/rswaty/bps_docs_parse/raw/main/all_bps_docs/10080_1_2_3_7.docx"

# Generate URLs by replacing the model_code in the URL template
urls <- sapply(unique_model_codes, function(code) {
  gsub("10080_1_2_3_7", paste0(code), url_template)
})

# Create clickable hyperlinks with BPS names, model codes, and REL_PERCENT without decimal places
clickable_names <- sapply(seq_along(urls), function(i) {
  paste0('<a href="', urls[i], '" target="_blank">', bps_urls$BPS_NAME[i], ' (', bps_urls$BPS_MODEL[i], ') - ', round(bps_urls$REL_PERCENT[i]), '%</a>')
})

# Render the list of hyperlinked BPS names as HTML
html_output <- HTML(paste(clickable_names, collapse = "<br>"))

# Print the HTML output
html_output



```


<br>




## Most Prevalent Existing Vegetation Types

```{r evt chart, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#| echo: false
evtname <- evt_data  |> 
  group_by(EVT_NAME) |> 
  summarize(ACRES = sum(ACRES),
            REL_PERCENT = sum(REL_PERCENT)) |> 
  arrange(desc(REL_PERCENT)) |> 
  filter(!EVT_NAME %in% c("Open Water","Developed-Roads")) |> 
  top_n(n = 10, wt = REL_PERCENT)

# plot
evtChart <-
  ggplot(data = evtname, aes(x = EVT_NAME, y = REL_PERCENT)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 Existing Vegetation Types",
    caption = "Data from landfire.gov",
    x = "",
    y = "Percent of landscape") +
  scale_x_discrete(limits = rev(evtname$EVT_NAME),
                   labels = function(x) str_wrap(x, width = 18)) +
  coord_flip() +
  theme_bw(base_size = 14)



evtChart
```

<br>

## Existing Vegetation Cover

The Existing Vegetation Cover (EVC) map is a visual representation of EVC classifications across the subregion. The chart below the map provides a breakdown of each vegetation cover classification and their relative distribution across the forest.

```{r evc chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
evcname <- evc_data %>% 
  mutate(type = if_else(Value %in% 11, "Open Water",
                                             if_else(Value %in% 12, "Snow / Ice",
                                                     if_else(Value %in% c(13:25), "Developed",
                                                             if_else(Value %in% 31, "Barren",
                                                                     if_else(Value %in% c(60:70), "Agriculture",
                                                                             if_else(Value %in% 32, "Quarries",
                                                                                     if_else(Value %in% 100, "Sparse Vegetation",
                                                                                             if_else(Value %in% c(101:199), "Tree",
                                                                                                     if_else(Value %in% c(201:299), "Shrub",
                                                                                                             if_else(Value %in% c(301:399), "Herb",
                                                                                                                     "Other")))))))))))



# create reverse substr() function
revSubstr <- function(x, start, stop) {
  x <- strsplit(x, "")
  sapply(x, 
         function(x) paste(rev(rev(x)[start:stop]), collapse = ""), 
         USE.NAMES = FALSE)  }

# create cover column based on 2nd and 3rd to last Values of classname
# if "Other" type, make 0
evcname <- evcname %>% mutate(cover = as.numeric(if_else(Value > 100,
                                                         revSubstr(evcname$CLASSNAMES, start = 2, stop = 3),
                                                         "0")))

# create bin breaks for grouping
breaks <- seq(0, 100, 10)
# create intervals for grouping and summarize
# also create factor order for "type"
evcgroup <- evcname %>%
  mutate(interval = cut(cover,
                        breaks, 
                        include.lowest = TRUE, 
                        right = T,
                        labels = c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", 
                                   "80-89", "90-100")),
         type = factor(type, levels = c("Tree", "Shrub", "Herb", "Open Water", "Snow / Ice", "Developed", "Agriculture", "Sparse Vegetation", "Barren", "Quarries", "Other"))) %>%
  group_by(type, interval) %>%
  summarize(COUNT = sum(Freq),
            acres = sum(ACRES),
            rel_percent = sum(REL_PERCENT))

# add label and legend names based on condition
evcgroup <- evcgroup %>% mutate(label = if_else(type %in% c("Tree", "Shrub", "Herb"),
                                                paste0(type, " Cover = ", interval, "%"), as.character(type)),
                                legend = if_else(type %in% c("Tree", "Shrub", "Herb", "Open Water"),
                                                 type, as.factor("Other")))

# turn current label order to factors
evclabel.list <- evcgroup$label
evcgroup <- evcgroup %>% mutate(label = fct_rev(factor(label, evclabel.list)))

# join in custom cols column to color bars by specific label

evc_group_cols <- read.csv("data/evc_group_acc.csv")

evcgroup <- left_join(evcgroup, evc_group_cols, by = "label")

evcgroup$label <- factor(evcgroup$label, levels = rev(evcgroup$label))

evcgroup <- evcgroup %>%
  filter(rel_percent > 0.01)

# plot
evcChart <-
  ggplot(data = evcgroup, aes(x = label, y = rel_percent, fill = colors)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Existing Vegetation Cover",
    caption = "Data from landfire.gov",
    x = "",
    y = "Percent of landscape") +
  scale_fill_identity() +
  coord_flip() +
  theme_classic(base_size = 12)+
  theme(legend.position = "none") 


evcChart

```

<br>

## Existing Vegetation Height

The Existing Vegetation Height (EVH) map showcases EVH across the forest. The chart below the map provides the percentage of the landscape represented by each EVH height.

```{r evh chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}

# load evh attribute table
evhname <- evh_data %>% 
  mutate(type = if_else(Value %in% 11, "Open Water",
                                             if_else(Value %in% 12, "Snow / Ice",
                                                     if_else(Value %in% c(13:25), "Developed",
                                                             if_else(Value %in% 31, "Barren",
                                                                     if_else(Value %in% c(60:70), "Agriculture",
                                                                             if_else(Value %in% 32, "Quarries",
                                                                                     if_else(Value %in% 100, "Sparse Vegetation",
                                                                                             if_else(Value %in% c(101:199), "Tree",
                                                                                                     if_else(Value %in% c(201:299), "Shrub",
                                                                                                             if_else(Value %in% c(301:399), "Herb",
                                                                                                                     "Other"))))))))))) %>%
  mutate(height_m = if_else(type %in% "Tree", (Value -100),
                            if_else(type %in% "Shrub", ((Value - 200) / 10),
                                    if_else(type %in% "Herb", ((Value - 300) / 10), 0))) %>%
           as.character() %>% as.numeric())

# create bin breaks for grouping
breaks <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90, 95, 100)

# create intervals for grouping and summarize
# also create factor order for "type"
evhgroup <- evhname %>%
  mutate(interval = cut(height_m,
                        breaks, 
                        include.lowest = TRUE, 
                        right = F,
                        labels = c("0", "0.1-0.2", "0.2-0.3", "0.3-0.4" ,"0.4-0.5", "0.5-0.6", "0.6-0.7", "0.7-0.8", "0.8-0.9", "0.9-1.0", "1-5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35", "35-40", "40-45", "45-50", "50-55", "55-60", "60-65", "65-70", "70-75", "75-80", "80-85", "85-90", "90-95", "95-100")),
         type = factor(type, levels = c("Tree", "Shrub", "Herb", "Open Water", "Snow / Ice", "Developed", "Agriculture", "Sparse Vegetation", "Barren", "Quarries", "Other"))) %>%
  group_by(type, interval) %>%
  summarise(COUNT = sum(Freq),
            acres = sum(ACRES),
            rel_percent = sum(REL_PERCENT))



# add label and legend names based on condition
evhgroup <- evhgroup %>% mutate(label = if_else(type %in% c("Tree", "Shrub", "Herb"),
                                                paste0(type, " Height = ", interval, " m"), as.character(type)),
                                legend = if_else(type %in% c("Tree", "Shrub", "Herb", "Open Water"),
                                                 type, as.factor("Other")))
# turn current label order to factors
evhlabel.list <- evhgroup$label
evhgroup <- evhgroup %>% mutate(label = fct_rev(factor(label, evhlabel.list)))


# join in custom cols column to color bars by specific label

evh_group_cols <- read.csv("data/evh_group_acc.csv")

evhgroup <- left_join(evhgroup, evh_group_cols, by = "label")

evhgroup$label <- factor(evhgroup$label, levels = rev(evhgroup$label))

evhgroup <- evhgroup %>%
  filter(rel_percent > 0.01)


# plot
evhChart <-
ggplot(data = evhgroup, aes(x = label, y = rel_percent, fill = colors)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Existing Vegetation Height",
    subtitle = "",
    caption = "\nData from landfire.gov.",
    x = "",
    y = "Percent of landscape") +
  scale_fill_identity() +
  coord_flip() +
  theme_bw()

evhChart



# plot with original color scheme
# evhChart <-
# ggplot(data = evhgroup, aes(x = label, y = REL_PERCENT, fill = legend)) +
#   geom_bar(stat = "identity") +
#   labs(
#     title = "Existing Vegetation Height",
#     subtitle = "landscape_name",
#     caption = "Data from landfire.gov.",
#     x = "",
#     y = "percent of landscape") +
#   scale_fill_manual(Values = cols, name = "") +
#   coord_flip() +
#   theme_bw()
# 
# evhChart





```
